\documentclass[11pt]{report}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{microtype}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  citecolor=blue,
  plainpages=false,
  pdftitle={Technical Report: Instance-Specific Portfolio Selection},
  pdfauthor={Christian Speck}
}

\title{Technical Report\\[0.4em]\large Instance-Specific Portfolio Selection for Algorithm Configuration}
\author{Christian Speck}
\date{\today}

\newtheorem{definition}{Definition}[chapter]
\newtheorem{remark}{Remark}[chapter]

\begin{document}

\hypersetup{pageanchor=false}
\maketitle
\pagenumbering{roman}
\tableofcontents
\clearpage
\hypersetup{pageanchor=true}
\pagenumbering{arabic}

\chapter{Introduction}

The central goal of this project is to study \emph{instance-specific parameter selection} under a restricted action space. We assume that a practitioner has already chosen a finite portfolio of parameter settings for a single underlying algorithm. At decision time, the system observes features of a problem instance and selects one portfolio member. The system does \emph{not} perform online continuous optimization of the parameters. Instead, it performs online \emph{routing} among a fixed set of candidate configurations.

This framing is closely aligned with the motivation of Instance-Specific Algorithm Configuration (ISAC): different regions of the instance space often favor different algorithmic behaviors, and observable features can be used to route a new instance to an appropriate configuration. The main difference in this repository is pedagogical and architectural. Rather than starting with a heavy benchmark and a full clustering pipeline, we begin with a minimal simulation whose structure matches the target decision problem.

\chapter{Problem Formulation}

\section{Instances, Features, and Portfolio Elements}

Let $\mathcal{X}$ denote the space of problem instances. For each instance $x \in \mathcal{X}$, we observe a feature vector
\[
\phi(x) \in \mathbb{R}^d,
\]
where $\phi$ is a feature extractor that is available at decision time.

Let the fixed portfolio of parameter settings be
\[
\Theta = \{\theta^{(1)}, \theta^{(2)}, \dots, \theta^{(K)}\},
\]
where each $\theta^{(k)} \in \mathbb{R}^p$ represents one preselected algorithm configuration. The key modeling choice is that $\Theta$ is finite and fixed before online deployment.

\begin{definition}
A portfolio selector is a mapping
\[
\pi : \mathbb{R}^d \to \{1,2,\dots,K\},
\]
which takes the observable feature vector $\phi(x)$ and returns the index of the portfolio element to execute.
\end{definition}

\section{Performance and Regret}

Let $c(x, \theta^{(k)})$ denote the cost of running configuration $\theta^{(k)}$ on instance $x$. Depending on the domain, this cost may be runtime, loss, error, or a weighted objective combining several metrics. We assume that lower values are better.

The oracle cost for instance $x$ is
\[
c^*(x) = \min_{1 \le k \le K} c(x, \theta^{(k)}).
\]

The \emph{instance regret} of choosing portfolio element $\theta^{(k)}$ on instance $x$ is
\[
r(x, \theta^{(k)}) = c(x, \theta^{(k)}) - c^*(x).
\]

Thus the expected regret of a selector $\pi$ under an instance distribution $\mathcal{D}$ is
\[
\mathbb{E}_{x \sim \mathcal{D}}
\left[
c(x, \theta^{(\pi(\phi(x)))}) - c^*(x)
\right].
\]

The learning objective is to construct $\pi$ so as to minimize this expected regret.

\section{Why This Is Not Reinforcement Learning}

The online decision in this project is a single-shot contextual routing problem. At each timestep, the system observes features and chooses a portfolio member. The environment does not require long-horizon credit assignment, policy improvement through repeated environmental interaction, or continuous online reconfiguration of the parameter vector.

Mathematically, this is much closer to contextual decision making or supervised cost-sensitive classification than to sequential reinforcement learning. The state is summarized by observable features, the action set is finite, and the per-instance loss is immediately measurable once the selected configuration has been evaluated.

\chapter{Synthetic Benchmark Design}

\section{Latent Regimes}

To make the problem nontrivial, the simulation assumes that instances arise from latent regimes
\[
z \in \{1,2,\dots,M\}.
\]
Each regime corresponds to a subset of the instance space for which different parameter settings are advantageous.

For regime $z$, we define:
\begin{itemize}[leftmargin=2em]
\item a feature-space center $\mu_z \in \mathbb{R}^d$,
\item an ideal parameter vector $\theta_z^* \in \mathbb{R}^p$.
\end{itemize}

An instance from regime $z$ is generated by sampling features according to
\[
\phi(x) = \mu_z + \varepsilon_x,
\quad
\varepsilon_x \sim \mathcal{N}(0, \sigma_x^2 I),
\]
followed by an ideal parameter vector
\[
\tilde{\theta}(x) = \operatorname{clip}\bigl(\theta_z^* + g(\phi(x)) + \varepsilon_\theta, 0, 1\bigr),
\quad
\varepsilon_\theta \sim \mathcal{N}(0, \sigma_\theta^2 I),
\]
where $g(\phi(x))$ introduces mild within-regime variation and $\operatorname{clip}$ enforces bounded parameter values.

\section{Cost Model}

Each portfolio element is a fixed vector $\theta^{(k)}$. The benchmark defines a base instance difficulty $b(x)$ and then measures mismatch between the chosen configuration and the latent ideal configuration:
\[
c(x, \theta^{(k)}) = b(x) + \lambda \lVert \theta^{(k)} - \tilde{\theta}(x) \rVert_2^2 + \eta,
\]
where $\lambda > 0$ is a mismatch penalty and $\eta$ is small observational noise.

This cost model yields three important properties:
\begin{enumerate}[leftmargin=2em]
\item Different regimes can favor different portfolio elements.
\item Features are predictive, but not perfectly predictive, of the best configuration.
\item A finite portfolio can be objectively evaluated by regret against the instance-wise best portfolio member.
\end{enumerate}

\section{Current Repository Instantiation}

The current implementation in the repository defines a compact synthetic portfolio benchmark with:
\begin{itemize}[leftmargin=2em]
\item three latent regimes,
\item a small number of observable features,
\item a two-dimensional parameter vector,
\item a four-element fixed portfolio containing conservative, aggressive, balanced, and robust parameter settings.
\end{itemize}

This benchmark lives in \texttt{src/isac/core/portfolio.py}. Its role is to provide a minimal experimental object with the same decision geometry as the intended real problem.

\chapter{Relation to ISAC}

\section{Conceptual Alignment}

ISAC begins from the hypothesis that feature-based structure in the instance space can be exploited to improve algorithm configuration. In broad terms, the logic is:
\begin{enumerate}[leftmargin=2em]
\item represent instances by feature vectors,
\item normalize or preprocess those features,
\item partition the instance space,
\item choose or learn a good configuration for each region,
\item route future instances according to their observed features.
\end{enumerate}

The benchmark in this repository captures the same core idea while simplifying the experimental substrate. Instead of immediately implementing the full offline clustering and per-cluster configuration optimization pipeline, we start from a benchmark where the true performance landscape is controllable and interpretable.

\section{A Minimal Selector Family}

The most direct selector family to study after the benchmark is implemented is a nearest-prototype rule. Suppose we compute centroids $\hat{\mu}_1, \dots, \hat{\mu}_M$ in feature space and estimate the empirically best portfolio member in each region, denoted $k_1^*, \dots, k_M^*$. Then a simple selector has the form
\[
\pi(\phi(x)) = k_{j(x)}^*,
\quad
j(x) = \arg\min_{1 \le m \le M} \lVert \phi(x) - \hat{\mu}_m \rVert_2^2.
\]

This style of rule is attractive because it is transparent, cheap at inference time, and closely related to the classical ISAC intuition.

\section{Baseline Selector Families}

To support systematic comparison, the repository evaluates a small set of baseline selector families under one shared train/test pipeline. In this report, a \emph{baseline} means a selector that is deliberately simple, easy to interpret, and valuable as a reference point even if it is not the most sophisticated model in the repository.

\paragraph{Clustering selector.}
The clustering-style baseline first partitions training instances into clusters and then assigns each cluster the empirically best portfolio member. If $a(i)$ denotes the cluster assignment of training instance $x_i$, then cluster $m$ receives label
\[
k_m^* = \arg\min_{1 \le k \le K}
\frac{1}{|\{i : a(i)=m\}|}
\sum_{i : a(i)=m} c(x_i, \theta^{(k)}).
\]
At inference time, a new instance is mapped to its nearest cluster center and inherits that cluster's portfolio choice.

\paragraph{Privileged classifier.}
The privileged classification comparator treats the oracle best portfolio index
\[
y_i = \arg\min_{1 \le k \le K} c(x_i, \theta^{(k)})
\]
as a class label and trains a model to predict $y_i$ from the feature vector $\phi(x_i)$. The current lightweight implementation uses a nearest-centroid classifier as a simple stand-in for richer discriminative models.

\paragraph{Regression selector.}
The regression-style baseline fits one predictor per portfolio element:
\[
\hat{c}_k(\phi(x)) \approx c(x, \theta^{(k)}),
\quad k = 1, \dots, K.
\]
The selected action is then
\[
\pi(\phi(x)) = \arg\min_{1 \le k \le K} \hat{c}_k(\phi(x)).
\]
This decomposition is useful because it predicts the whole performance table rather than only the best action. It is a strong baseline because it aligns closely with the actual decision problem: estimate each candidate's cost, then choose the cheapest one.

\section{Learned Selector Families}

Beyond the core baselines, the repository also studies richer learned selectors whose purpose is not merely to provide a floor for comparison, but to test whether additional representation learning, nonlinear discrimination, or temporally aware soft routing can outperform the simpler references.

\paragraph{Multi-layer perceptron (MLP) selector.}
A more expressive discriminative model is a feed-forward multi-layer perceptron (MLP)
\[
f_\omega : \mathbb{R}^d \to \mathbb{R}^K
\]
that maps instance features to one logit per portfolio element. The selected action is
\[
\pi(\phi(x)) = \arg\max_{1 \le k \le K} f_\omega(\phi(x))_k.
\]
This model family is important because it can represent nonlinear decision boundaries in feature space while still remaining a direct predictor of the oracle best portfolio member.

\paragraph{Deep graph clustering for algorithm configuration (DGCAC)-inspired selector.}
A natural modern extension is to learn a representation before clustering. Let $e_\psi(\phi(x)) \in \mathbb{R}^q$ denote a learned embedding of the observable feature vector. The routing pipeline becomes
\[
\phi(x) \xrightarrow{\;e_\psi\;} z(x) \in \mathbb{R}^q
\xrightarrow{\;\text{cluster}\;} m(x)
\xrightarrow{\;\text{cluster label}\;} \pi(\phi(x)).
\]
This mirrors the architectural idea in recent deep-clustering approaches to instance-specific configuration: representation learning and routing are coupled, rather than assuming that the raw feature space is the right geometry for clustering. The current repository includes a lightweight approximation of this idea by using a low-dimensional learned embedding followed by cluster-wise portfolio assignment.
In its present form, the embedding stage is trained as a denoising autoencoder and the routing stage uses soft latent-space assignments, so the selector computes an expected runtime across nearby clusters rather than committing to a single hard cluster label.

\chapter{Evaluation Quantities}

\section{Average Cost and Regret}

For a dataset $\{x_i\}_{i=1}^n$, the empirical average cost of a selector $\pi$ is
\[
\hat{C}(\pi) = \frac{1}{n} \sum_{i=1}^n c\bigl(x_i, \theta^{(\pi(\phi(x_i)))}\bigr).
\]
The empirical average regret is
\[
\hat{R}(\pi) = \frac{1}{n} \sum_{i=1}^n
\left(
c\bigl(x_i, \theta^{(\pi(\phi(x_i)))}\bigr) -
\min_{1 \le k \le K} c(x_i, \theta^{(k)})
\right).
\]

\section{Global Fallback Baseline}

A natural baseline is the best \emph{instance-oblivious} portfolio member:
\[
k_{\mathrm{global}}^* =
\arg\min_{1 \le k \le K}
\frac{1}{n} \sum_{i=1}^n c(x_i, \theta^{(k)}).
\]

This baseline is important because it answers a concrete question: how much value do the features provide beyond a single globally fixed parameter setting?

\section{Train/Test Comparison Protocol}

The current repository uses a simple but important evaluation discipline. A selector is fit on one synthetic sample from the benchmark distribution and evaluated on a separate held-out sample. Feature normalization parameters are estimated only on the training split and then reused on the test split. This keeps the comparison between clustering, classification, and regression selectors honest and prevents test-time information leakage through normalization.

\section{Next Research Approximation}

The first modern paper targeted by the repository is the deep-clustering approach of Song et al.\ on instance-specific algorithm configuration. The full paper uses graph-based instance representations and unsupervised deep clustering. The repository does not yet include graph encoders, but it now includes a deep graph clustering for algorithm configuration (DGCAC)-inspired selector that preserves the core structural idea:
\begin{enumerate}[leftmargin=2em]
\item learn a compact latent representation,
\item cluster instances in the latent space,
\item aggregate cluster-conditioned runtime estimates to choose a portfolio element.
\end{enumerate}

This staged approximation is deliberate. It keeps the benchmark lightweight while still moving the selector family in the direction suggested by contemporary ISAC research.

\section{Non-Baseline Selector Models}

The repository now contains several selectors that are more than simple benchmark baselines. These models are the main research objects in the codebase because they attempt to learn richer decision structure than nearest-centroid classification, plain regression, or hard cluster lookup. The purpose of this section is to describe their internal logic in enough detail that the reader can understand what each model is trying to capture and why one might expect it to outperform simpler baselines under the right benchmark conditions.

\subsection{MLP Selector}

The MLP selector is the repository's main direct nonlinear discriminative model. Its basic role is to learn a map
\[
f_\omega : \mathbb{R}^d \to \mathbb{R}^{N_s},
\]
where $d$ is the observable feature dimension and $N_s$ is the selector's learned portfolio size cap. Each output coordinate corresponds to one learned parameter prototype in that selector's own portfolio. The final action is chosen by
\[
\pi(\phi(x)) = \arg\max_{1 \le s \le N_s} f_\omega(\phi(x))_s.
\]

The important modeling choice is that the MLP does not try to model the full runtime table explicitly. Instead, it learns a direct partition of feature space into action regions. This makes it a useful stress test for whether the benchmark can be solved by a powerful but still fairly generic feed-forward discriminative model.

In the current implementation, the MLP selector uses PyTorch, validation-based model selection, and a cost-sensitive training criterion so that the optimization target is closer to downstream regret than to pure classification accuracy. Conceptually, the model is trying to answer the question
\begin{quote}
given the current observable state, which learned parameter prototype should be selected?
\end{quote}
This is attractive when the best-action surface is highly nonlinear but still recoverable from current observations alone. It is less attractive when the problem becomes strongly history-dependent or when switching costs matter enough that a purely local decision rule becomes myopic.

\subsection{DGCAC-Inspired Selector}

The DGCAC-inspired selector is the repository's main representation-learning model. It is motivated by deep-clustering approaches to instance-specific algorithm configuration, especially settings in which the raw observed feature space is not assumed to be the right geometry for routing. The selector first learns an embedding
\[
z_t = e_\psi(h_t) \in \mathbb{R}^q,
\]
where $h_t$ is a feature vector or short temporal context summary and $q$ is the latent dimension. The latent representation is then used for soft cluster assignment. If the cluster centers are $m_1,\dots,m_M$, then the routing weights take the form
\[
w_j(z_t) \propto \exp\!\left(-\frac{\lVert z_t - m_j\rVert_2^2}{\tau}\right),
\]
with temperature parameter $\tau > 0$ controlling how soft the assignment is.

These routing weights are used to aggregate cluster-conditioned runtime estimates, so the final action is chosen by minimizing an expected predicted cost rather than committing immediately to a single cluster label. In prose, the selector tries to do three things at once:
\begin{enumerate}[leftmargin=2em]
\item denoise and compress the observation into a latent state,
\item place that latent state near semantically similar training states,
\item use nearby clusters to infer which portfolio element should be preferred.
\end{enumerate}

In the current repository, the model is still a lightweight approximation rather than a full reproduction of graph-based DGCAC. It uses feature-vector inputs rather than graph encoders, and its embedding stage is trained with a denoising-style objective together with decision-oriented heads. Even so, it already captures the main intuition behind modern deep ISAC systems: routing quality depends on representation quality.

\subsection{Temporal Soft Cluster ISAC}

Temporal Soft Cluster ISAC is currently the most ambitious non-baseline selector in the repository and the strongest model in the enriched dynamic benchmark. Its design starts from a simple observation: in the dynamic simulator, the correct action depends not only on the current partially observed feature vector, but also on recent history, latent drift, and the penalty for switching actions too aggressively.

Accordingly, the model does not cluster the raw current observation alone. Instead, it builds a temporalized state summary
\[
g_t = \Gamma(o_{t-k+1}, \dots, o_t),
\]
where $o_t$ is the current observation and $\Gamma$ is a handcrafted temporal featurization that includes current observation terms, running summaries, and short-horizon trend information. These temporal descriptors are then softly clustered rather than hard-clustered. If $u_1,\dots,u_M$ denote temporal cluster prototypes, the selector computes membership weights
\[
\alpha_j(g_t) \propto \exp\!\left(-\frac{\lVert g_t - u_j\rVert_2^2}{\tau}\right).
\]

Each cluster is equipped with a local runtime model or local performance summary over the selector's learned portfolio. The model then forms a weighted predicted cost for each candidate parameter set,
\[
\hat{c}_s(g_t) = \sum_{j=1}^M \alpha_j(g_t)\,\hat{c}_{j,s}(g_t),
\]
and selects the action that minimizes this quantity after including a switching-aware adjustment. If the previous action was $a_{t-1}$, the deployed decision rule is conceptually
\[
a_t =
\arg\min_s
\left(
\hat{c}_s(g_t) + \lambda_{\mathrm{switch}}\mathbf{1}\{s \neq a_{t-1}\}
\right).
\]

This model is important because it is the first selector in the repository that explicitly combines all of the following:
\begin{itemize}[leftmargin=2em]
\item temporal context,
\item soft routing rather than hard assignment,
\item local specialization,
\item switching-aware action choice.
\end{itemize}

Those ingredients align closely with the current dynamic simulator, which includes partial observability, regime changes, multimodal performance structure, and explicit switching costs. The model's recent performance improvements suggest that the combination of temporal context and soft local specialization is materially more useful than either ingredient alone.

\section{Separability Diagnostics}

Before sophisticated methods are introduced, the benchmark should support basic diagnostics:
\begin{itemize}[leftmargin=2em]
\item the proportion of instances on which each portfolio element is optimal,
\item cluster or regime confusion under simple selectors,
\item regret reduction relative to the global baseline,
\item sensitivity to feature noise and overlap.
\end{itemize}

\section{Three-Noise Robustness Sweep}

To evaluate a selector across all three difficulty sources simultaneously, the repository uses a Cartesian sweep over feature noise, parameter noise, and runtime noise. Let
\[
\Sigma_f,\ \Sigma_p,\ \Sigma_r
\]
denote the chosen grids for these three sources. For every triple
\[
(\sigma_f, \sigma_p, \sigma_r) \in \Sigma_f \times \Sigma_p \times \Sigma_r,
\]
the benchmark samples a fresh train/test split, fits each selector, and records held-out metrics.

If $R(\pi; \sigma_f, \sigma_p, \sigma_r, s)$ denotes the average regret of selector $\pi$ under random seed $s$, then the sweep reports quantities such as
\[
\bar{R}(\pi) =
\frac{1}{|\Sigma_f||\Sigma_p||\Sigma_r||S|}
\sum_{\sigma_f \in \Sigma_f}
\sum_{\sigma_p \in \Sigma_p}
\sum_{\sigma_r \in \Sigma_r}
\sum_{s \in S}
R(\pi; \sigma_f, \sigma_p, \sigma_r, s),
\]
as well as a worst-case summary
\[
R_{\max}(\pi) =
\max_{\sigma_f,\sigma_p,\sigma_r,s}
R(\pi; \sigma_f, \sigma_p, \sigma_r, s).
\]

These two summaries serve different purposes: the former measures average quality over the grid, while the latter measures robustness to difficult combinations of overlapping regimes, intra-regime parameter drift, and stochastic runtime perturbations.

\chapter{Current Results}

\section{Living Results Policy}

This chapter is intended to be updated whenever the repository's current benchmark results change in a meaningful way. In practice, that means a report update should accompany any commit that changes selector logic, evaluation protocol, or the precomputed sweep artifact used by the dashboard.

\section{Latest Validated Snapshots}

The repository recently switched from a fixed shared action set to \emph{selector-specific learned portfolios} with a configurable cap $N_s$ and current default $N_s = 12$. Because of that protocol change, the most important numbers to trust are the fresh post-change snapshots validated directly against the current code path.

The privileged classifier should still be interpreted as a truth-aware comparator rather than a realistic deployment model, since it relies on oracle best-config labels during training.

\subsection{Static Representative Snapshot}

For the static benchmark, the latest validated run uses
\[
\sigma_f = 0.40,\qquad \sigma_p = 0.08,\qquad \sigma_r = 0.03,
\]
with 360 training instances, 360 test instances, and seed 7. Under this learned-portfolio evaluation, the ranking is:

\begin{center}
\begin{tabular}{lccc}
\toprule
Selector & Average regret & Average runtime & Accuracy \\
\midrule
Cluster ISAC & 0.044573 & 2.002495 & 0.575000 \\
MLP Selector & 0.056062 & 2.013985 & 0.375000 \\
DGCAC-inspired & 0.068389 & 2.026312 & 0.294444 \\
Regressor & 0.072619 & 2.030542 & 0.275000 \\
Privileged Classifier & 0.076731 & 2.034653 & 0.261111 \\
\bottomrule
\end{tabular}
\end{center}

The main surprise is that Cluster ISAC remains the strongest method on regret under the representative static setting even after the learned-portfolio cap is increased to $N_s = 12$. This keeps the static benchmark firmly in the regime where region-specific routing still matters.

\subsection{Dynamic Enriched Snapshot}

For the dynamic benchmark, the latest validated run uses horizon 16 together with
\[
\texttt{drift\_scale}=0.22,\quad
\texttt{regime\_switch\_prob}=0.18,\quad
\texttt{switching\_cost}=0.04,
\]
\[
\texttt{observation\_noise}=0.10,\quad
\texttt{missing\_feature\_prob}=0.22,\quad
\texttt{multimodal\_surface\_scale}=0.30,
\]
with 80 training episodes, 80 test episodes, and seed 11. Measured by average total penalty (regret plus switching cost), the ranking is:

\begin{center}
\begin{tabular}{lcccc}
\toprule
Selector & Avg. regret & Avg. switch cost & Avg. total penalty & Accuracy \\
\midrule
Temporal Soft Cluster ISAC & 0.061812 & 0.007031 & 0.068843 & 0.317188 \\
Regressor & 0.064847 & 0.012531 & 0.077378 & 0.350000 \\
Cluster ISAC & 0.063036 & 0.014687 & 0.077723 & 0.324219 \\
DGCAC-inspired & 0.071707 & 0.006781 & 0.078488 & 0.281250 \\
MLP Selector & 0.067181 & 0.015219 & 0.082399 & 0.357031 \\
Privileged Classifier & 0.078971 & 0.013656 & 0.092627 & 0.318750 \\
\bottomrule
\end{tabular}
\end{center}

This dynamic result is consistent with the current simulator design: once partial observability, drift, switching costs, and multimodal quality surfaces are all active, the problem rewards \emph{history-aware soft routing}. The new temporal soft-cluster model improves on both the plain regressor and the original hard Cluster ISAC by reducing switching cost without sacrificing local specialization.

\subsection{Concrete Algorithmic Benchmark}

The repository now also includes a more faithful algorithm-configuration benchmark in which selectors do not compete against an abstract latent runtime surface. Instead, they configure a concrete adaptive forecasting algorithm with three tunable parameters:
\[
\alpha \quad \text{(level adaptation)}, \qquad
\beta \quad \text{(trend adaptation)}, \qquad
g \quad \text{(volatility gate)}.
\]
In this setting, an episode is a drifting observed stream generated from latent regimes with different trend, volatility, mean reversion, and shock behavior. Offline portfolio construction is performed by optimizing this concrete algorithm on diverse training episodes, collecting strong incumbents, and compressing them into a capped portfolio.

This benchmark matters because it better matches the intended project goal: given observed feature data, learn a small set of strong parameter settings for one algorithm and then infer which setting to use online. The repository currently supports two offline portfolio-construction modes for this benchmark:
\begin{itemize}[leftmargin=2em]
\item a lightweight random-search builder for fast local experimentation,
\item an optional SMAC3-driven builder for optimizer-based offline portfolio discovery.
\end{itemize}

The resulting selector evaluation path is now structurally closer to real ISAC usage than the original synthetic distance-to-ideal benchmark, since the portfolio is derived from actual algorithm optimization rather than only from latent ideal parameter vectors.

\paragraph{SMAC-backed representative snapshot.}
With a smaller but fully optimizer-backed run using horizon 12, four training episodes, four test episodes, and 12 SMAC trials per episode, the current ranking on average regret is:

\begin{center}
\begin{tabular}{lccc}
\toprule
Selector & Portfolio source & Portfolio size & Average regret \\
\midrule
Temporal Soft Cluster ISAC & SMAC & 4 & 0.006280 \\
DGCAC-inspired & SMAC & 4 & 0.008318 \\
Cluster ISAC & SMAC & 4 & 0.010955 \\
Privileged Classifier & SMAC & 4 & 0.012048 \\
Regressor & SMAC & 4 & 0.012764 \\
MLP Selector & SMAC & 4 & 0.018369 \\
\bottomrule
\end{tabular}
\end{center}

Even though this is still a modest benchmark snapshot rather than a large offline study, it is already informative: once the portfolio comes from real optimizer-discovered incumbents rather than purely synthetic prototype derivation, the temporal soft-cluster selector remains the strongest method among the tested selectors, and the improved DGCAC-inspired selector now emerges as the second-best method on this concrete algorithmic task. This is encouraging because it suggests that the selector's advantage is not purely an artifact of the older latent-ideal benchmark formulation.

\section{Current Takeaway}

At the moment, the benchmark supports two different short-run conclusions:
\begin{itemize}[leftmargin=2em]
\item in the static learned-portfolio setting, Cluster ISAC is the current regret leader,
\item in the enriched dynamic setting, Temporal Soft Cluster ISAC is now the strongest model on total penalty,
\item in the concrete SMAC-backed algorithmic benchmark, Temporal Soft Cluster ISAC remains the strongest selector and DGCAC-inspired is now the next best model,
\item the simpler regressors and centroid-style baselines are still useful references, but they are no longer the strongest models once the problem is framed as offline optimizer-driven portfolio construction plus online routing,
\item MLP remains a meaningful neural baseline, but it is not yet competitive with the strongest structured selectors on the current algorithmic setup,
\item the privileged classifier remains useful as a supervision upper-bound reference, but it is not the realistic deployment winner.
\end{itemize}

This is a useful outcome for the project. It suggests that allowing each selector to derive its own parameter portfolio materially changes the comparison, and it reinforces the value of evaluating selectors on a concrete algorithm rather than only on an abstract latent runtime surface. The present results also sharpen the design lesson: structured temporal routing with portfolio awareness is more promising than pure label prediction once the offline portfolio is built from real parameter search.

\chapter{Path to Real Data}

\section{ASlib as the Most Natural Follow-Up}

The most natural real-data path is the Algorithm Selection Library (ASlib). ASlib standardizes scenarios that provide instance features and per-instance performance tables over a finite set of candidate solvers. This is extremely close to the desired project structure.

For the purposes of this repository, one may reinterpret each candidate solver in an ASlib scenario as a \emph{portfolio element}. In some settings that portfolio element is literally a different algorithm. In others, it can be viewed as a stand-in for a precomputed parameterization of one base method. The important structural point is the same: a finite action set is available offline, and the online task is to select among those actions from features.

\section{OpenML-Derived Scenarios}

OpenML-derived scenarios are particularly appealing because they more closely resemble machine-learning model and hyperparameter selection at the dataset level. These scenarios are likely to be a good bridge between the synthetic benchmark and the eventual target application in which one has dataset characteristics and must choose among several fixed configurations.

\chapter{Implementation Notes}

\section{Repository Role of the Report}

This report is intended to function as a living technical chapter. Whenever the repository gains a new benchmark, selector, dataset loader, or evaluation protocol, the report should be updated so that:
\begin{itemize}[leftmargin=2em]
\item the notation remains consistent,
\item the mathematical objective is explicit,
\item the experimental abstractions are documented in prose rather than only in code.
\end{itemize}

\section{Build Process}

The PDF version of this report is generated from this \LaTeX{} source using \texttt{latexmk}. The repository includes a small build script so that the command
\begin{verbatim}
./scripts/build_report.sh
\end{verbatim}
produces an updated PDF in \texttt{docs/report/build/technical\_report.pdf}.

\section{PyTorch Migration}

The learning-oriented selector families are now implemented with PyTorch while preserving the repository's original NumPy-facing \texttt{fit}/\texttt{predict} interface. This change keeps the experiment harness stable while making it easier to introduce modern neural baselines, reuse optimizers such as Adam, and extend the DGCAC-inspired selector toward deeper encoders. In practical terms, the migration currently affects:
\begin{itemize}[leftmargin=2em]
\item the linear regression selector, via tensor-based least-squares solves,
\item the MLP selector, via supervised cross-entropy training,
\item the DGCAC-inspired selector, via a denoising autoencoder and soft latent routing.
\end{itemize}

This is a useful architectural boundary: benchmark generation remains lightweight and deterministic, while model training can now scale toward more realistic representation-learning experiments.

\section{Interactive Dashboard}

To support rapid iteration on the benchmark design, the repository also includes a local dashboard implemented with Streamlit and Plotly. Its purpose is not merely presentational. It acts as an inspection tool for checking whether the benchmark exhibits the right qualitative structure:
\begin{itemize}[leftmargin=2em]
\item separable but overlapping feature-space regimes,
\item a nontrivial action mix across portfolio members,
\item measurable regret differences between routing heuristics,
\item interpretable instance-level failures.
\end{itemize}

The dashboard exposes the sampled feature geometry, selector-induced parameter surfaces, robustness curves over the three noise variables, and aggregate summaries such as worst-case regret and win rate over the sweep grid. This is useful because benchmark design errors often reveal themselves visually before they reveal themselves through aggregate metrics alone.

\chapter{Conclusion}

The project begins with a deliberately narrow question: can observable instance characteristics guide the online selection of one configuration from a finite, precomputed parameter portfolio? That question is already rich enough to support meaningful mathematical definitions, interpretable simulations, and rigorous evaluation. By establishing that foundation first, the repository can later expand toward richer ISAC-inspired models without losing clarity about the underlying decision problem.

\end{document}
